Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-20 PAPER.tex
-went through whole paper for typos and small changes

We form $100$ bootstraps, each comprised of an array of length $N$ that is created by a random resampling of the original data, with replacement. The standard deviation over the $100$ bootstraps gives an error estimate for our dataset. As shown in Figure \ref{fig:toy_error1} by the grey star, the error computed from bootstrapping matches our theoretical prediction.

@@

Going back to our toy model, we apply a sliding boxcar average to $10$ samples at a time, thus reducing the number of independent samples to $N/10 = 100$. Bootstrapping this time-averaged noise, using the same method as described earlier (drawing $N=1000$ samples), under-estimates the error by a factor of $\sim3$. This occurs because we are drawing more samples than independent ones available, and thus some samples are repeated multiple times in all bootstraps, leading to less variation between the bootstraps. In fact, the error derived from bootstrapping is a strong function of the number of samples that are drawn (Figure \ref{fig:toy_error1}, black points), and we can both under-estimate the error by drawing too many or over-estimate it by drawing too few. However, if we know that we have $100$ independent samples, the error associated with drawing $100$ samples with replacement does match the theoretical prediction as expected.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/toy_error1.png}
	\caption{Error estimation produced by bootstrapping as a function of the number of samples drawn when sampling with replacement. The star represents the error associated with drawing $1000$ samples from a length $1000$ array of a Gaussian random signal. The black points correspond to time-averaged data (correlated data) which has $100$ independent samples. They illustrate how errors can be underestimated if drawing more samples than there are independent samples in the data. The estimated errors match up with the theoretical prediction only at $N=100$.}
	\label{fig:toy_error1}
\end{figure}

@@

Therefore, we use a slightly modified bootstrapping method. For each bootstrap, we draw $N-1=4$ samples without replacement and draw the last slot randomly with replacement. In doing so there is a small chance a bootstrap will consist of $5$ independent samples, but even with $4$ our sensitivity is nearly maximized. One may wonder whether this change is still a legitimate way of error estimating since random sampling only occurs for one value in a bootstrap. However, as long as the number of possible variations is greater than the number of bootstraps we perform, it is a valid way to uncover the inherent variability in a dataset.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/toy_error2.png}
	\caption{The fraction of independent elements in one bootstrap sample as a function of the total number of elements in a dataset (i.e. $4$ out of $5$ elements are independent would yield a fraction of $0.8$). A fraction of $1.0$ means that all elements in the bootstrap sample are independent, and therefore sensitivity is maximized (alternately, this axis can be thought of as power spectrum sensitivity, where sensitivity increases moving upwards along the y-axis). Two bootstrapping methods are shown here --- sampling all elements with replacement (grey) and sampling only the last element with replacement (black). The former fails to maximize sensitivity since some elements end up being repeated due to random sampling. The `outlier' points lying at a fraction of $1.0$ occur in the off-chance that the last spot in the sample happens to be the one missing element instead of a repeated element --- this is more likely to happen with a small total number of elements but also results in a more dramatically different fraction as compared to the case of a repeated element in the last spot, hence why these points appear like outliers.}
	\label{fig:toy_error2}
\end{figure}

@@

Because foreground spectra are smooth in frequency, they preferentially show up at low delay, or $k$ modes. For a particular baseline length, there is a maximum delay imposed on foregrounds, which corresponds to the light-crossing time between the two antennas in the baseline. For longer baselines, this value increases, producing what is known as ``the wedge" (\citealt{parsons_et_al2012b}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{vedantham_et_al2012}; \citealt{thyagarajan_et_al2013}; \citealt{pober_et_al2013}; \citealt{datta_et_al2010}). The wedge describes a region in $k$-space contaminated by foregrounds, bounded by baseline length (which is proportional to $k_{\perp}$) and delay (which is proportional to $k_{\parallel}$). Properties of the wedge can be used to isolate and remove foregrounds, as done by P64, \citet{parsons_et_al2014}, and \citet{jacobs_et_al2015}.

@@

\begin{itemize}
\item \textbf{Null Test}: A null test is a type of jackknife test that removes the astronomical signal from data in order to investigate underlying systematics. Several null tests on data are described in \citet{keating_et_al2016}. For example, one can divide data into two subsets by separating odd and even julian dates, or the first half of the observing season from the second. Subtracting the two removes signal that is common to both subsets, including foregrounds and EoR. The resulting power spectrum should be consistent with thermal noise estimates; if it is not, it suggests the presence of a systematic that differs from one of the data subsets to the other (i.e. doesn't get subtracted perfectly). 
\item \textbf{Traditional Jackknife}: In a more broad sense, it is important to perform many jackknife tests in order to instill confidence in a final result, since a stable result should be consistent through all of them. A result must be steadfast throughout all jackknives no matter how the data is sliced. Jackknives can be taken along several different axes --- for example, one could start with a full dataset, and compute a power spectrum each time as a day of data is removed, or a baseline is removed. This type of jackknife would reveal bias present only at certain LSTs (such as a foreground source), for example, or misbehaving baselines.
\end{itemize}

@@

We form $3$ different power spectrum estimates shown in Figure \ref{fig:toy_bias1}. The first is a null test where we subtract $\textbf{x}_{2}$ from $\textbf{x}_{1}$ for both measurements $a$ and $b$. This is equivalent to splitting up a full dataset along an axis (in this case, time) and subtracting the two to remove sky signal that should ideally be present in both. We cross-multiply measurements $a$ and $b$ to form an un-biased (noise-wise) estimate (blue curve). The second estimate, shown in red, is the same null test with the foreground systematic removed (eliminate $\textbf{fg}$ in Equations \ref{eq:bias1} and \ref{eq:bias2}). Finally, we also show the noise power spectrum (green).

From this test we see a clear difference between the null test with the presence of the foregrounds, and the power spectrum of noise. This signifies a non-EoR bias that is only present in either $\textbf{x}_{1}$ or $\textbf{x}_{2}$.

@@

For our analysis, we use $8$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (P64 uses an RA range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ MHz ($20$ channels), identical to the analysis in P64. We note that the PAPER-64 dataset that we use in this case study differs from that in P64 mainly by the fringe-rate filter. \cc{Danny says the next part is wrong, but that's what's in Zaki's paper... } In P64, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ cm signal visibilities. Here we use the same method used in P64, where mock gaussian noise visibilities are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One specific change from P64 is that we add this simulated signal into the chain before the final fringe-rate filter is applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased estimate of signal loss, likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the same level if different $21$ cm simulations were used remains an open question.}

@@

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average, and this negative cross-term power arises from inverting the covariance matrix $\textbf{C}_{r}$. Mathematically,

@@

\noindent where the data-only term (first term) and EoR-only term (last term) are approximately diagonal matrices. The middle two terms, however, have a larger amount of power in their off-diagonal terms relative to the diagonal elements, since $\textbf{e}$ and $\textbf{x}$ are statistically unrelated. Upon inversion, these off-diagonal terms become negative, driving the cross-terms in Equation \ref{eq:pout_expand} to have negative power. This effect is explained in more depth in \citet{switzer_et_al2015}.

@@

There is a second component of scatter that is less intuitive, only applies (at a low level) to the weighted case, and represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values. We correct for this as explained in the next section.

@@

Two additional horizontal lines are shown in Figure \ref{fig:sigloss_modeloop} which denote power spectrum values, post-signal loss correction, for two other regularization schemes. We multiply an identity matrix element-wise to $\hat{\textbf{C}}$ (i.e. inverse variance weighting, shown in red), and also add an arbitrary level of $\textbf{I}$ to $\hat{\textbf{C}}$ (cyan). We see that all three regularization schemes (black, red, cyan) perform similarly.

@@

Re-analysis of the PAPER-64 data included a detailed study using several independently generated noise simulations. What we found was that these simulations all agreed but were discrepant with the previous analytic sensitivity calculations. The analytic calculation is only an approximation, however the differences were large enough (factors of $10$ in some cases) to warrant a careful investigation. The analytic calculation attempts to combine a large number of pieces of information in an approximate way, and when re-considering some of the approximations we have found there to be large effects. What follows here is an accounting of the differences which have been discovered.

The sensitivity prediction \cc{need refs} for a power spectral analysis of interferometric $21$ cm data, in temperature-units, is:

@@

\begin{itemize}
\item $X^{2}Y$: Conversion factors from observing coordinates (angles on the sky) to cosmological coordinates (co-moving distances). For $z=8.4$, $X^{2}Y = 5 \times 10^{11} h^{-3} Mpc^{3} str^{-1} GHz^{-1}$.
\item $\Omega_{eff}$: The effective primary beam area in steradians \footnote{\cc{ref equation}}. The effective beam area changes with the application of a fringe-rate filter, since parts of the beam are up-weighted and down-weighted. Using numbers from Table 1 in \citet{parsons_et_al2016}, $\Omega_{eff} = 0.74^{2}/0.24$ for an optimal fringe-rate filter. 
\item $T_{sys}$: The system temperature is set by:

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $200$ K, yielding $T_{sys} = 487$ K at $150$ MHz. This is lower than in P64 because of previous failure to correct for using a median statistic.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The revised limit in \cc{cite kolopanis paper} uses three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of $N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is $N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}}$, where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines entering the group ($N_{bls}/N_{gps}$) and then as square root of the number of cross multiplied pairs ($\sqrt{N_{gps}^{2} - N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

Our revised sensitivity estimate for PAPER-64 is shown in comparison with that of P64 in Figure \ref{fig:sense_check}. Together, the revised parameters yield a decrease in sensitivity (higher noise floor) by a factor of $\sim7$. This is consistent with

@@

\noindent where $\Delta\nu$ is channel spacing, $\Delta t$ is integration time, $N_{days}$ is the number of daily counts for a particular time and frequency that went into our LST binned set, and $N_{pols}$ is the number of polarizations ($2$ for Stokes I). This RMS temperature sets the variance of the Gaussian random noise.

@@

As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised of independent samples. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power spectrum that is a function of $k$ and time. In P64, a second round of bootstrapping occurred over the baseline bootstrap and time axes simultaneously. Random values were sampled with replacement along both axes, drawing as many values as there are number of time integrations. Final power spectrum limits were then computed by taking the mean and standard deviation over this second bootstrap axis (which is typically large, on the order of $N_{boot} \sim 400$). We emphasize that in this previous version, the number of elements sampled per bootstrap (number of times $\sim$ $700$) greatly exceeds the number of independent LST samples, under-estimating errors.

Because our dataset is fringe-rate filtered, we have substantially increased our integration time, resulting in fewer independent samples (from $\sim700$ samples to $8$ samples). A random draw of $700$ samples from this dataset therefore has many repeated samples, and the variance between hundreds ($N_{boot}$) of these random samples is smaller than the true underlying variance of the data.

@@

Power spectrum estimates with these bootstrapping changes for PAPER-64 fringe-rate filtered data are shown in Figure \ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As shown in the figure, when the baseline average includes many repeated samples the error is over-estimated, while when only one sample in the same average is allowed the possibility of being a repeat, the error on the noise-dominated higher $k$ modes converges to the theoretical noise level.  However, if more elements are drawn for each bootstrap than the number of independent samples (by drawing many times along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits below the predicted noise level.

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/data_errors.png}
	\caption{$2\sigma$ power spectrum errors (not upper limits) using PAPER-64 data and three different bootstrapping methods. The data is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (unweighted) is used in order to disentangle the effects of bootstrapping from signal loss. Three different bootstrapping methods are shown: the blue and black use methods where bootstrapping occurs on the baseline axis only (not the time axis), and they differ by how the baselines are randomly sampled. It is evident that sensitivity is maximized when baselines are sampled in a way that ensures mostly independent samples (only the last baseline spot is sampled randomly), as shown by the black curve. In red, bootstrapping occurs over both the baseline and time axes, illustrating how errors can be under-estimated if sampling more values than independent ones (fringe-rate filtering reduces the number of independent samples).}
	\label{fig:data_errors}
\end{figure}

@@

In Section \ref{sec:BiasOverview} we highlighted some common sources of bias that can show up as a power spectrum detection and imitate EoR. We discussed the importance of using jackknife and null tests for instilling confidence in an EoR detection, as well as identifying other sources of biases. Here we demonstrate methods used by PAPER-64 to mitigate foreground and noise bias and we perform jackknife and null tests in order to characterize the stability and implications of our results.

@@

The highest sensitivity power spectrum result using PAPER-64 data and all revised methods, shown in Figure \ref{fig:ps1_data}, has positive biases at low $k$ values. As discussed in Section \ref{sec:BiasTypes}, these detections are most likely attributable to foreground leakage. Here we demonstrate $3$ null tests performed on PAPER-64 data that verify that the positive detections are indeed due to foreground variation and not attributable to EoR.

>>>>

We form $100$ bootstraps, each comprised of an array of length $N$ that is created by a random resampling of the original data, with replacement. The standard deviation over the $100$ bootstraps gives an error estimate for our dataset. As shown in Figure \ref{fig:toy_error1} by the grey star, the error computed from bootstrapping matches our theoretical prediction (green).

@@

Going back to our toy model, we apply a sliding boxcar average to $10$ samples at a time, thus reducing the number of independent data samples to $N/10 = 100$. Bootstrapping this time-averaged noise, using the same method as described earlier (drawing $N=1000$ elements per bootstrap sample), under-estimates the error by a factor of $\sim3$. This occurs because we are drawing more samples than independent ones available, and thus some samples are repeated multiple times in all bootstraps, leading to less variation between the bootstraps. In fact, the error derived from bootstrapping is a strong function of the number of elements that are drawn per bootstrap (Figure \ref{fig:toy_error1}, black points), and we can both under-estimate the error by drawing too many or over-estimate it by drawing too few. However, if we know that we have $100$ independent samples, the error associated with drawing $100$ samples with replacement does match the theoretical prediction as expected.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/toy_error1.png}
	\caption{Error estimation produced by bootstrapping as a function of the number of elements drawn per bootstrap when sampling with replacement. The star represents the error associated with drawing $1000$ elements from a length $1000$ array of a Gaussian random signal. The black points correspond to time-averaged data (correlated data) which has $100$ independent samples. They illustrate how errors can be underestimated if drawing more elements than there are independent samples in the data. The estimated errors match up with the theoretical prediction only at $N=100$.}
	\label{fig:toy_error1}
\end{figure}

@@

Therefore, we use a slightly modified bootstrapping method. For each bootstrap, we draw $N-1=4$ samples without replacement and draw the last slot randomly with replacement. In doing so there is a small chance a bootstrap will consist of $5$ independent measurements, but even with $4$ our sensitivity is nearly maximized. One may wonder whether this change is still a legitimate way of error estimating since random sampling only occurs for one value in a bootstrap. However, as long as the number of possible variations is greater than the number of bootstraps we perform, it is a valid way to uncover the inherent variability in a dataset.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/toy_error2.png}
	\caption{The fraction of independent elements in one bootstrap sample as a function of the total number of elements in a dataset (i.e. if $4$ out of $5$ elements are independent, this would yield a fraction of $0.8$). A fraction of $1.0$ means that all elements in the bootstrap sample are independent, and therefore sensitivity is maximized (alternately, this axis can be thought of as power spectrum sensitivity, where sensitivity increases moving upwards along the y-axis). Two bootstrapping methods are shown here --- sampling all elements with replacement (grey) and sampling only the last element with replacement (black). The former fails to maximize sensitivity since some elements end up being repeated due to random sampling. The `outlier' points lying at a fraction of $1.0$ at the far left of the plot occur in the off-chance that the last spot in the sample happens to be the one missing element instead of a repeated element --- this is more likely to happen with a small total number of elements but also results in a more dramatically different fraction as compared to the case of a repeated element in the last spot, hence why these points appear like outliers.}
	\label{fig:toy_error2}
\end{figure}

@@

Because foreground spectra are smooth in frequency, they preferentially show up at low delay, or $k$ modes. For a particular baseline length, there is a maximum delay imposed on foregrounds, which corresponds to the light-crossing time between two antennas in a baseline. For longer baselines, this value increases, producing what is known as ``the wedge" (\citealt{parsons_et_al2012b}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{vedantham_et_al2012}; \citealt{thyagarajan_et_al2013}; \citealt{pober_et_al2013}; \citealt{datta_et_al2010}). The wedge describes a region in $k$-space contaminated by foregrounds, bounded by baseline length (which is proportional to $k_{\perp}$) and delay (which is proportional to $k_{\parallel}$). Properties of the wedge can be used to isolate and remove foregrounds, as done by P64, \citet{parsons_et_al2014}, and \citet{jacobs_et_al2015}.

@@

\begin{itemize}
\item \textbf{Null Test}: A null test is a type of jackknife test that removes the astronomical signal from data in order to investigate underlying systematics. Several null tests on data are described in \citet{keating_et_al2016}. For example, one can divide data into two subsets by separating odd and even julian dates, or the first half of the observing season from the second. Subtracting the two removes signal that is common to both subsets, including foregrounds and EoR. The resulting power spectrum should be consistent with thermal noise estimates; if it is not, it suggests the presence of a systematic that differs from one of the data subsets to the other (i.e. doesn't get subtracted perfectly). 
\item \textbf{Traditional Jackknife}: In a more broad sense, it is important to perform many jackknife tests in order to instill confidence in a final result. A stable result must be steadfast throughout all jackknives no matter how the data is sliced. Jackknives can be taken along several different axes --- for example, one could start with a full dataset, and compute a new power spectrum every time as a day of data is removed, or a baseline is removed. This type of jackknife would reveal bias present only at certain LSTs (such as a foreground source), for example, or misbehaving baselines.
\end{itemize}

@@

We form $3$ different power spectrum estimates shown in Figure \ref{fig:toy_bias1}. The first is a null test where we subtract $\textbf{x}_{2}$ from $\textbf{x}_{1}$ for both measurements $a$ and $b$. This is equivalent to splitting up a full dataset along an axis (in this case, time) and subtracting the two to remove sky signal that should ideally be present in both. We cross-multiply measurements $a$ and $b$ to form an un-biased (thermal noise-wise) estimate (blue curve). The second estimate, shown in red, is the same null test with the foreground systematic removed (eliminate $\textbf{fg}$ in Equations \ref{eq:bias1} and \ref{eq:bias2}). Finally, we also show the noise power spectrum (green).

From this test we see a clear difference between the null test with the presence of the foregrounds, and the power spectrum of noise. This signifies a non-EoR bias that is only present in either $\textbf{x}_{1}$ or $\textbf{x}_{2}$, but not both.

@@

For our analysis, we use $8$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (P64 uses a slightly different RA range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ MHz ($20$ channels), identical to the analysis in P64. We note that the PAPER-64 dataset that we use in this case study differs from that in P64 mainly by the fringe-rate filter. \cc{Danny says the next part is wrong, but that's what's in Zaki's paper... } In P64, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ cm signal visibilities. Here we use the same method used in P64, where mock gaussian noise visibilities (mock EoR signals) are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.  \footnote{One specific change from P64 is that we add this simulated signal into the chain before the final fringe-rate filter is applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased estimate of signal loss, likely due to the use of the fringe-rate filter as a simulator. Whether the loss would be estimated at the same level if different $21$ cm simulations were used remains an open question.}

@@

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average, and this negative cross-term power arises from inverting the covariance matrix $\textbf{C}_{r}$. Mathematically, the covariance of the data plus EoR is:

@@

\noindent where the data-only term (first term) and EoR-only term (last term) are approximately diagonal matrices. The middle two terms, however, have a larger amount of power in their off-diagonal terms relative to their diagonal elements, since $\textbf{e}$ and $\textbf{x}$ are statistically unrelated. Upon inversion, these off-diagonal terms become negative, driving the cross-terms in Equation \ref{eq:pout_expand} to have negative power. This effect is explained in more depth in \citet{switzer_et_al2015}.

@@

There is a second component of scatter that is less intuitive, only applies (at a low level) to the weighted case, and represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values. We correct for this, as explained later in this section.

@@

In addition to our thermal noise prediction (green), two additional horizontal lines are shown in Figure \ref{fig:sigloss_modeloop} which denote power spectrum values, post-signal loss correction, for two other regularization schemes. We multiply an identity matrix element-wise to $\hat{\textbf{C}}$ (i.e. inverse variance weighting, shown in red), and also add an arbitrarily chosen level of $\textbf{I}$ to $\hat{\textbf{C}}$ (cyan). We see that all three regularization schemes (black, red, cyan) perform similarly at their best.

@@

Re-analysis of the PAPER-64 data included a detailed study using several independently generated noise simulations. What we found was that these simulations all agreed but were discrepant with the previous analytic sensitivity calculations. The analytic calculation is only an approximation, however the differences were large enough (factors of $10$ in some cases) to warrant a careful investigation. The analytic calculation attempts to combine a large number of pieces of information in an approximate way, and when re-considering some of the approximations, we have found there to be large effects. What follows here is an accounting of the differences which have been discovered.

The sensitivity prediction (\citealt{parsons_et_al2012a}) for a power spectral analysis of interferometric $21$ cm data, in temperature-units, is:

@@

\begin{itemize}
\item $X^{2}Y$: Conversion factors from observing coordinates (angles on the sky) to cosmological coordinates (co-moving distances). For $z=8.4$, $X^{2}Y = 5 \times 10^{11} h^{-3} Mpc^{3} str^{-1} GHz^{-1}$.
\item $\Omega_{eff}$: The effective primary beam area in steradians (\citealt{parsons_et_al2010}, \citealt{pober_et_al2012}). The effective beam area changes with the application of a fringe-rate filter, since parts of the beam are up-weighted and down-weighted. Using numbers from Table 1 in \citet{parsons_et_al2016}, $\Omega_{eff} = 0.74^{2}/0.24$ for an optimal fringe-rate filter. 
\item $T_{sys}$: The system temperature is set by:

@@

where $\nu$ are frequencies in GHz. We use a receiver temperature of $200$ K, yielding $T_{sys} = 487$ K at $150$ MHz. This is lower than in P64 because of previous failure to correct for using a median statistic.
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from taking the real part of the power spectrum estimates after squaring visibility measurements. In P64, a factor of $2$ was mistakenly used.
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$. The updated limits in \cc{cite kolopanis paper} use three separation types.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling the original integration time using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is the fringe-rate filtered integration time, $w$ is the fringe-rate profile, and the integral is taken over all fringe-rates. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In P64, this number was set to $135$. However, because we divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, because our LST coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of $N_{days}$ as:
\begin{equation}
 \frac{1}{N_{days}} = \sqrt{\Big\langle\frac{1}{N_{i}^{2}} \Big\rangle_{i}},
 \end{equation}
\noindent where $i$ is over LST (\citealt{jacobs_et_al2015}). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In P64, this number was the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects the fact that baselines are averaged into groups before cross-multiplying data. Our revised estimate for the parameter is:
\begin{equation}
N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}},
\end{equation}
\noindent where, for the P64 analysis, $N_{gps} = 5$. Each baseline group averages down linearly as the number of baselines entering the group ($N_{bls}/N_{gps}$) and then as the square root of the number of cross-multiplied pairs ($\sqrt{N_{gps}^{2} - N_{gps}}$). For the revised PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

@@

Our revised sensitivity estimate for PAPER-64 is shown in comparison with that of P64 in Figure \ref{fig:sense_check}. Together, the revised parameters yield a decrease in sensitivity (higher noise floor) by a factor of $\sim7$.

@@

\noindent where $\Delta\nu$ is channel spacing, $\Delta t$ is integration time, $N_{days}$ is the number of daily counts for a particular time and frequency that went into our LST-binned set, and $N_{pols}$ is the number of polarizations ($2$ for Stokes I). This RMS temperature sets the variance of the Gaussian random noise.

@@

As discussed in Section \ref{sec:ErrorOverview}, bootstrapping is only a valid way of estimating errors if a dataset is comprised of independent samples. The PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power spectrum that is a function of $k$ and time. In P64, a second round of bootstrapping occurred over the baseline bootstrap and time axes simultaneously. Random values were sampled with replacement along both axes, drawing as many values as there are number of time integrations. Final power spectrum limits were then computed by taking the mean and standard deviation over this second bootstrap axis (which is typically large, on the order of $N_{boot} \sim 400$). We emphasize that in this previous version, the number of elements sampled per bootstrap ($N \sim $ number of times $\sim$ $700$) greatly exceeds the number of independent LST samples ($\sim$$8$), under-estimating errors.

Because our dataset is fringe-rate filtered, we have substantially increased our integration time, resulting in fewer independent samples (from $\sim700$ samples to $8$ samples). A random draw of $700$ measurements from this dataset therefore has many repeated values, and the variance between hundreds ($N_{boot}$) of these random samples is smaller than the true underlying variance of the data.

@@

Power spectrum estimates with these bootstrapping changes for PAPER-64 fringe-rate filtered data are shown in Figure \ref{fig:data_errors}. The estimates are unweighted in order to disentangle the effects of bootstrapping from signal loss. As shown in the figure, when the baseline average includes many repeated samples the error is over-estimated, while when only one sample in the same average is allowed the possibility of being a repeat, the error on the noise-dominated higher $k$ modes converges to the theoretical noise level.  However, if more elements are drawn for each bootstrap than the number of independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits below the predicted noise level.

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},width=\columnwidth]{plots/data_errors.png}
	\caption{$2\sigma$ power spectrum errors (not upper limits) using PAPER-64 data and three different bootstrapping methods. The data is fringe-rate filtered and a weighting matrix of $\textbf{I}$ (unweighted) is used in order to disentangle the effects of bootstrapping from signal loss. Three different bootstrapping methods are shown: the blue and black use methods where bootstrapping occurs on the baseline axis only (not the time axis), and they differ by how the baselines are randomly sampled. It is evident that sensitivity is maximized when baselines are sampled in a way that ensures mostly independent samples (only the last baseline spot is sampled randomly), as shown by the black curve. In red, bootstrapping occurs over both the baseline and time axes, illustrating how errors can be under-estimated if sampling more values than independent ones (fringe-rate filtering reduces the number of independent samples). We use the method illustrated by the black curve in our updated analysis, which is shown to agree with the analytic prediction for noise (green).}
	\label{fig:data_errors}
\end{figure}

@@

In Section \ref{sec:BiasOverview} we highlighted some common sources of bias that can show up as power spectrum detections and imitate EoR. We discussed the importance of using jackknife and null tests for instilling confidence in an EoR detection, as well as identifying other sources of biases. Here we demonstrate methods used by PAPER-64 to mitigate foreground and noise bias and we perform null tests in order to characterize the stability and implications of our results.

@@

The highest sensitivity power spectrum result for PAPER-64 using the updated analysis presented in this paper, shown in Figure \ref{fig:ps1_data}, has positive biases at low $k$ values. As discussed in Section \ref{sec:BiasTypes}, these detections are most likely attributable to foreground leakage. Here we demonstrate $3$ null tests performed on PAPER-64 data that verify that the positive detections are indeed due to foreground variation and not attributable to EoR.
